Mutual Modeling in Human-Robot Interaction - Scientific part

### 2014-03-20 project-mutual-modeling.tex
Using tex macro to typeset models

@@

    \item We could continue the list of what is X is $\mathcal{M}(A,B,X)$: beliefs,
        emotions, history, status, …

@@

A partner model is probably not a "box", \ie not a monolithical representation
but rather a mosaic of information fragments about the partner, with various
grain sizes and various life cycles. This mosaic is elaborated through various
mechanisms, first for building an initial model of the partner and then for
updating this model.  As two students meet for the first time, mutual models are
initialized by the assumptions they make upon each other from cues such as
his/her membership to large categories (age, culture, profession, ...) include
stereotypes (sportsmen, junkie, business women, Swiss,…) as well as physical
appearance. Scholars studied how initial modeling impacts communication. In
their experiments on initial MM, Slugoski, Lalljee, Lamb \& Ginsburg (1993)
pretended to their subjects that their (fake) partner had or had not received
the same information. They observed that the subjects adapted their dialogue by
focusing the explanation on the items that (s)he was supposed to ignore. Brennan
(1991) showed that the subjects used different initial strategies in forming
queries depending on who they were told their partner was.

Initial common grounds are also initiated by co-presence: they include events to
which A and B attended together (Clark \& Marshall, 1981) in the physical space
or in our cultural space (\eg ‘09-11’). Co-presence means that we can refer to
a shared objects and event,-s it does of course not imply that we give them the
same meaning. Namely, a shared screen does not mean a shared understanding
(Dillenbourg \& Traum, 2006).

@@

Mutual modeling does not occur in a vacuum but it is highly contextualized. REF
reviews the features of the collaborative situation, namely the media
(cotemporality… ), may facilitate or hamper mutual modeling. Hutchins (SILENCE)
reported a study in which a short silence between two pilots was perfectly
interpreted because it occurred in a highly constrained communication context.
Some environments are more productive than others in helping peers to detect
their misunderstandings. Roschelle (REF) reformulate the design of CSCL
interfaces as providing ways for peer to detect and repairs their
misunderstanding. In CSCW, researchers have explored various so-called
"awareness tools", \ie functionalities that inform A is about B's actions that
A could not directly perceived because B was working on a different subset of
the virtual space. Different awareness tools will be addressed in this
contribution since they allowed us to manipulate experimentally the MM activity.

@@

@@

@@

To report experiments on MM, we use a simple notational system: "A knows that B
knows X", noted $\mathcal{M}(A,B,X)$ is a simplistic but hopefully useful reduction to
reason on mutual modeling. This notation does not mean A as an explicit,
monolithic representation of B: It is an abstraction referring to complex
socio-cognitive processes.

We refer to the degree of accuracy of the mutual model as
$\mathcal{M}^{\circ}(A,B,X)$. To estimate, the MM effort, we need 3 variables.

@@

    \item Tasks vary a lot with respect to how much they require.  The grounding
        criterion – denoted $\mathcal{M}^{\circ}_{min}$  - refers to how
        important it is to mutually share a piece of information X to succeed
        the task T.. It can be computed as the probability to succeed T despite
        the fact X is not grounded. In the next studies, we estimate
        $\mathcal{M}^{\circ}_{min}(A,B,X)$ as the correlation between
        $\mathcal{M}^{\circ}(A,B,X)$ and task performance.

@@

    \item The cost of grounding X refers to the physical and cognitive effort
        necessary to perform a grounding act α: a verbal repair (\eg
        rephrasing), a deictic gesture, a physical move to adopt one partner's
        viewpoint, etc. This cost varies according to media features ( Clark \&
        Brennan (REF….).

@@

Based on these 3 parameters, the probability of making an action $\alpha_{t1}$ about
content X at time $t1$ during task T for increasing $\mathcal{M}^{\circ}(A,B,X)$
at time $t2$ is the ratio between how much it is needed  ((2)-(1)) and how much it
costs (3)(Traum \& Dillenbourg, 1996):

@@

\begin{itemize}
    \item $\mathcal{M}^{\circ}(A,B,X)$: Our experiments address different contents that can be
        represented in mutual models:

        \begin{enumerate}
            \item $\mathcal{M}(A,B,emotion)$: how accurately A perceives B's emotional state
                (study 3);

            \item $\mathcal{M}(A,B,actions)$ is about how well A guesses what action B has
                done (study 2) or will do next  (study 1)

            \item $\mathcal{M} (A,B,knowledge)$: how accurately A estimates B's knowledge
                with respect to the material they learn together (study 4 and 5)

@@

    \item $\mathcal{M}^{\circ\circ}(A,B,X,T)$ Our studies concern various collaborative tasks:
        argumentation (study 3), games (study 1 and 2) and concept mapping
        (study 4 and 5). By varying the tasks, we do actually vary the grounding
        criterion but they all require a "reasonably high" grounding criterion.
        All tasks are about building a solution or a representation together; we
        did not address everyday conversations.

@@

Measuring $\mathcal{M}^{\circ}(A,B,X)$ is methodologically difficult. We
discriminate two steps, first to capture $\mathcal{M}(A,B,X)$ and then to
estimate $\mathcal{M}^{\circ}(A,B,X)$.

@@

    \item Estimating $\mathcal{M}^{\circ}(A,B,X)$. The accuracy of
        $\mathcal{M}(A,B,X)$ can be
        estimated in 2 ways.
        \begin{itemize}
            \item Subjective accuracy: In study 3, we compute
                $\mathcal{M}^{\circ}(A,B,emotions)$ by measuring if A describe
                B's emotions in the same way B reports her emotions (
                ($\mathcal{M}(A,B)=\mathcal{M}(B,B)$)

            \item Objective accuracy: In studies 4 and 5, w we compute
                $\mathcal{M}^{\circ}(A,B,knowledge)$ by comparing
                $\mathcal{M}^{\circ}A,B,K)$ to B's actual knowledge as it has
                measured by a test.

@@

Our main issue is the symmetry questions: what is the relationship (fig.1)
between $\mathcal{M}^{\circ}(A,B,X)$ and $\mathcal{M}^{\circ}(B,A,X)$? A low
symmetry would mean that mutual modeling
is mainly an individual attitude/aptitude ($\mathcal{H}_{1}$). A high
correlation might support $\mathcal{H}_{2}$ since there is a low probability
that randomly formed pairs integrate peers with the same level of MM skills.

@@

\draw(0,0) node[anchor=north] (A) {\scriptsize A};
\draw(10,0) node[anchor=north] (B) {\scriptsize B};
\draw(5,2) node[anchor=north] {\scriptsize $\mathcal{M}^{\circ}(A, B, X)$};
\draw(5,-2) node[anchor=north] {\scriptsize $\mathcal{M}^{\circ}(B, A, X)$};
\draw[dashed, <->] (5,1) -- (5,-1.9) node[midway, sloped, above] {$\Delta_1$};
\draw[->] (A) to[bend left] (B);
\draw[->] (B) to[bend left] (A);

@@

\label{mm_symetry}
\end{figure*}

With triads, we may compute the accuracy of 6 models:
$\mathcal{M}^{\circ}(A,B)$, $\mathcal{M}^{\circ}(B,A)$,
$\mathcal{M}^{\circ}(A,C)$, $\mathcal{M}^{\circ}(C,A)$,
$\mathcal{M}^{\circ}(C,B)$ and $\mathcal{M}^{\circ}(B,C)$. This enables two
triangle questions (see figure 2):

Do A and B have the same accuracy when modeling C? $\Delta2 =
\Delta(\mathcal{M}(A,C,X), \mathcal{M}(B,C,X))$ If it is the same,
$\mathcal{H}_{2}$ and $\mathcal{H}_{3}$ gain over $\mathcal{H}_{1}$

Does C model more accurately A than B? $\Delta3= \Delta(\mathcal{M}(C,A,X),
\mathcal{M}(C,B,X))$ A positive answer would support $\mathcal{H}_{3}$.

@@

    \draw(5,2) node {\scriptsize $\mathcal{M}^{\circ}(A, C, X)$};
    \draw(5,-2) node {\scriptsize $\mathcal{M}^{\circ}(B, C, X)$};
    \draw[dashed, <->] (3,1.2) -- (3,-1.2) node[midway, sloped, above] {$\Delta_2$};
    \draw[->] (A) to (C);
    \draw[->] (B) to (C);

@@

    \draw(5,2) node {\scriptsize $\mathcal{M}^{\circ}(C, A, X)$};
    \draw(5,-2) node {\scriptsize $\mathcal{M}^{\circ}(C, B, X)$};
    \draw[dashed, <->] (3,1.2) -- (3,-1.2) node[midway, sloped, above]
    {$\Delta_3$};
    \draw[<-] (A) to (C);
    \draw[<-] (B) to (C);

@@

\label{mm_triangle}
\end{figure}

In addition, the comparison between ($\Delta2$) and ($\Delta3$) could tell us
whether the accuracy of mutual modelling depends more upon the modeller's effort
or the modellee's behaviour.

The rectangle questions

We could go further by comparing self- versus other modeling ($\Delta4$ in Fig.
3) as an indication of metacognitive skills. We could also see if modeling
skills depend upon what aspects are being modeled (X or Y), which would explain
vertical differences ($\Delta5$ in Fig. 3). We do not further develop these
questions because we don’t have the necessary data in our studies.

@@

    \draw(0,0) node (a) {\scriptsize $\mathcal{M}^{\circ}(A, B, X)$};
    \draw(10,0) node (b) {\scriptsize $\mathcal{M}^{\circ}(A, A, X)$};
    \draw(10,-4) node (c) {\scriptsize $\mathcal{M}^{\circ}(A, A, Y)$};
    \draw(0,-4) node (d) {\scriptsize $\mathcal{M}^{\circ}(A, B, Y)$};
    \draw[<->] (a) -- (b) node[midway, below] {$\Delta_4$};
    \draw[<->] (b) -- (c);
    \draw[<->] (c) -- (d);
    \draw[<->] (d) -- (a) node[midway, sloped, above] {$\Delta_5$};

>>>>

\newcommand{\M}[3]{{\mathcal{M}(#1, #2, #3)}}
\newcommand{\model}[3]{{$\mathcal{M}(#1, #2, #3)$}}
\newcommand{\Model}[3]{{$\mathcal{M}^{\circ}(#1, #2, #3)$}}

\newcommand{\concept}[1]{{\small \texttt{#1}}}
\newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}

@@

    \item We could continue the list of what is X is \model{A}{B}{X}: beliefs,
        emotions, history, status, …

@@

A partner model is probably not a "box", \ie not a monolithical representation
but rather a mosaic of information fragments about the partner, with various
grain sizes and various life cycles. This mosaic is elaborated through various
mechanisms, first for building an initial model of the partner and then for
updating this model.  As two students meet for the first time, mutual models are
initialized by the assumptions they make upon each other from cues such as
his/her membership to large categories (age, culture, profession, ...) include
stereotypes (sportsmen, junkie, business women, Swiss,...) as well as physical
appearance. Scholars studied how initial modeling impacts communication. In
their experiments on initial MM, Slugoski, Lalljee, Lamb \&
Ginsburg~\cite{slugoski1993attribution} pretended to their subjects that their
(fake) partner had or had not received the same information. They observed that
the subjects adapted their dialogue by focusing the explanation on the items
that (s)he was supposed to ignore. Brennan (1991) showed that the subjects used
different initial strategies in forming queries depending on who they were told
their partner was.

Initial common grounds are also initiated by co-presence: they include events to
which A and B attended together~\cite{clark2002definite} in the physical space
or in our cultural space (\eg ‘09-11’). Co-presence means that we can refer to
a shared objects and event,-s it does of course not imply that we give them the
same meaning. Namely, a shared screen does not mean a shared
understanding~\cite{dillenbourg2006sharing}.

@@

Mutual modeling does not occur in a vacuum but it is highly contextualized. REF
reviews the features of the collaborative situation, namely the media
(cotemporality… ), may facilitate or hamper mutual modeling. Hutchins (SILENCE)
reported a study in which a short silence between two pilots was perfectly
interpreted because it occurred in a highly constrained communication context.
Some environments are more productive than others in helping peers to detect
their misunderstandings. Roschelle~\cite{roschelle1995construction} reformulate
the design of CSCL interfaces as providing ways for peer to detect and repairs
their misunderstanding. In CSCW, researchers have explored various so-called
"awareness tools", \ie functionalities that inform A is about B's actions that A
could not directly perceived because B was working on a different subset of the
virtual space. Different awareness tools will be addressed in this contribution
since they allowed us to manipulate experimentally the MM activity.

@@

\paragraph{Building a model of a human during the interaction} Perspective
Taking is a human ability which allows one to put him/herself in another
person's point of view. Studied in psychology
literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
interacting with people by allowing one to reason on others' understanding of
the world in terms of visual perception, spatial descriptions, affordances and
beliefs, etc.  In the last years these notions have been gradually employed in
HRI.~\cite{Breazeal2006} presents a learning algorithm that takes into account
information about a teacher's visual perspective in order to learn a task.
~\cite{Johnson2005} apply visual perspective taking for action recognition
between two robots.~\cite{Trafton2005} use both visual and spatial perspective
taking to find out the referent indicated by a human partner.

The cognitive model that the robot builds for the agent it interacts with is
still simple and mostly focused on geometric features (who sees what? What are
our relative positions? etc.). Extending this knowledge with more subtle
perceptions (emotional state for instance) remains to be done.

\paragraph{Joint action}
Several theories dealing with collaboration~\cite{Cohen1991,Grosz1996,Clark1996}
emphasize that collaborative tasks have specific requirements compared to
individual ones, \eg, since the robot and the person share a common goal, they
have to agree on the manner to realize it, they must show their commitment to
the goal during execution, etc. Several robotic systems have already been built
based on these theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they all
have shown benefits of this approach. They have also shown how difficult it is
to manage turn-taking between communication partners and to interleave task
realization and communication in a generic way. Finally, today only few
systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into account at all
levels.

@@

Our knowledge base implements such a mechanism: when the robots infers a new
agent has been introduced in the knowledge base, it initializes a new,
independent, ontology for this agent. All the ontologies that are created share
the same common-sense knowledge, but rely on each agent's perspective for the
actual instantiation: the robot (geometrically) computes that the book is in
its own field of view, but not in the human one. The robot knowledge contains
the fact \stmt{book isVisible true} while the human model contains \stmt{book
isVisible false}.

Using separate knowledge models in the knowledge base, we have been able to
replicate the \emph{Sally and Ann} false-belief experiment with our
robots~\cite{warnier2012when}.

In our implementation, perspective taking is tightly connected to the symbolic
knowledge models, and since our knowledge base allows for storage of one
knowledge model per agent, we have been able to endow the robot with a simple
theory of mind: we explicitly model
what the robot knows about its partners in a symbolic way. This knowledge is
then re-used in different places, to correctly interpret what the human says,
or to plan tasks that are actually doable for the human.

@@

To report experiments on MM, we use a simple notational system: "A knows that B
knows X", noted \model{A}{B}{X} is a simplistic but hopefully useful reduction to
reason on mutual modeling. This notation does not mean A as an explicit,
monolithic representation of B: It is an abstraction referring to complex
socio-cognitive processes.

We refer to the degree of accuracy of the mutual model as
\Model{A}{B}{X}. To estimate, the MM effort, we need 3 variables.

@@

    \item Tasks vary a lot with respect to how much they require.  The grounding
        criterion – denoted $\mathcal{M}^{\circ}_{min}$  - refers to how
        important it is to mutually share a piece of information X to succeed
        the task T.. It can be computed as the probability to succeed T despite
        the fact X is not grounded. In the next studies, we estimate
        $\mathcal{M}^{\circ}_{min}(A,B,X)$ as the correlation between
        \Model{A}{B}{X} and task performance.

@@

    \item The cost of grounding X refers to the physical and cognitive effort
        necessary to perform a grounding act α: a verbal repair (\eg
        rephrasing), a deictic gesture, a physical move to adopt one partner's
        viewpoint, etc. This cost varies according to media features ( Clark \&
        Brennan (REF)).

@@

Based on these 3 parameters, the probability of making an action $\alpha_{t1}$ about
content X at time $t1$ during task T for increasing \Model{A}{B}{X}
at time $t2$ is the ratio between how much it is needed  ((2)-(1)) and how much it
costs (3)(Traum \& Dillenbourg, 1996):

@@

\begin{itemize}
    \item \Model{A}{B}{X}: Our experiments address different contents that can be
        represented in mutual models:

        \begin{enumerate}
            \item \model{A}{B}{emotion}: how accurately A perceives B's emotional state
                (study 3);

            \item \model{A}{B}{actions} is about how well A guesses what action B has
                done (study 2) or will do next  (study 1)

            \item \model{A}{B}{knowledge}: how accurately A estimates B's knowledge
                with respect to the material they learn together (study 4 and 5)

@@

    \item \Model{A,B}{X}{T} Our studies concern various collaborative tasks:
        argumentation (study 3), games (study 1 and 2) and concept mapping
        (study 4 and 5). By varying the tasks, we do actually vary the grounding
        criterion but they all require a "reasonably high" grounding criterion.
        All tasks are about building a solution or a representation together; we
        did not address everyday conversations.

@@

Measuring \Model{A}{B}{X} is methodologically difficult. We
discriminate two steps, first to capture \model{A}{B}{X} and then to
estimate \Model{A}{B}{X}.

@@

    \item Estimating \Model{A}{B}{X}. The accuracy of
        \model{A}{B}{X} can be
        estimated in 2 ways.
        \begin{itemize}
            \item Subjective accuracy: In study 3, we compute
                \Model{A}{B}{emotions} by measuring if A describe
                B's emotions in the same way B reports her emotions 
                ($\mathcal{M}(A,B)=\mathcal{M}(B,B)$)

            \item Objective accuracy: In studies 4 and 5, w we compute
                \Model{A}{B}{knowledge} by comparing
                $\mathcal{M}^{\circ}A,B,K)$ to B's actual knowledge as it has
                measured by a test.

@@

Our main issue is the symmetry questions: what is the relationship
(Fig.~\ref{mm_symmetry}) between \Model{A}{B}{X} and
\Model{B}{A}{X}? A low symmetry would mean that mutual modeling is
mainly an individual attitude/aptitude ($\mathcal{H}_{1}$). A high correlation
might support $\mathcal{H}_{2}$ since there is a low probability that randomly
formed pairs integrate peers with the same level of MM skills.

@@

\draw(0,0) node[anchor=north] (A) {\scriptsize A};
\draw(10,0) node[anchor=north] (B) {\scriptsize B};
\draw(5,2) node[anchor=north] {\scriptsize \Model{A}{ B}{ X}};
\draw(5,-2) node[anchor=north] {\scriptsize \Model{B}{ A}{ X}};
\draw[dashed, <->] (5,1) -- (5,-1.9) node[midway, sloped, above] {$\Delta_1$};
\draw[->] (A) to[bend left] (B);
\draw[->] (B) to[bend left] (A);

@@

\label{mm_symmetry}
\end{figure*}

With triads, we may compute the accuracy of 6 models:
\Model{A}{B}{X}, \Model{B}{A}{X}, \Model{A}{C}{X}, \Model{C}{A}{X},
\Model{C}{B}{X} and \Model{B}{C}{X}. This enables two
triangle questions (see Fig.~\ref{mm_triangles}):

Do A and B have the same accuracy when modeling C? $\Delta_2 =
\Delta(\M{A}{C}{X}, \M{B}{C}{X})$ If it is the same,
$\mathcal{H}_{2}$ and $\mathcal{H}_{3}$ gain over $\mathcal{H}_{1}$

Does C model more accurately A than B? $\Delta_3= \Delta(\M{C}{A}{X},
\M{C}{B}{X})$ A positive answer would support $\mathcal{H}_{3}$.

@@

    \draw(5,2) node {\scriptsize \Model{A}{ C}{ X}};
    \draw(5,-2) node {\scriptsize \Model{B}{ C}{ X}};
    \draw[dashed, <->] (3,1.2) -- (3,-1.2) node[midway, sloped, above] {$\Delta_2$};
    \draw[->] (A) to (C);
    \draw[->] (B) to (C);

@@

    \draw(5,2) node {\scriptsize \Model{C}{ A}{ X}};
    \draw(5,-2) node {\scriptsize \Model{C}{ B}{ X}};
    \draw[dashed, <->] (3,1.2) -- (3,-1.2) node[midway, sloped, above]
    {$\Delta_3$};
    \draw[<-] (A) to (C);
    \draw[<-] (B) to (C);

@@

\label{mm_triangles}
\end{figure}

In addition, the comparison between $\Delta_2$ and $\Delta_3$ could tell us
whether the accuracy of mutual modelling depends more upon the modeller's effort
or the modellee's behaviour.

\paragraph{The rectangle questions}

We could go further by comparing self- versus other modeling $\Delta_4$ in
Fig.~\ref{mm_rectangle}) as an indication of metacognitive skills. We could also
see if modeling skills depend upon what aspects are being modeled (X or Y),
which would explain vertical differences ($\Delta_5$ in
Fig.~\ref{mm_rectangle}). We do not further develop these questions because we
don’t have the necessary data in our studies.

@@

    \draw(0,0) node (a) {\scriptsize \Model{A}{ B}{ X}};
    \draw(10,0) node (b) {\scriptsize \Model{A}{ A}{ X}};
    \draw(10,-4) node (c) {\scriptsize \Model{A}{ A}{ Y}};
    \draw(0,-4) node (d) {\scriptsize \Model{A}{ B}{ Y}};
    \draw[<->] (a) -- (b) node[midway, below] {$\Delta_4$};
    \draw[<->] (b) -- (c);
    \draw[<->] (c) -- (d);
    \draw[<->] (d) -- (a) node[midway, sloped, above] {$\Delta_5$};
